### 2020-12-17 manuscript.tex
manuscript tidyups

\clearpage
\subsection*{Branch rate parameterisations}
\label{sect:rateparams}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/rateparameterisation.pdf}
\caption{\textbf{Branch rate parameterisations.} Top left: the prior density of a branch rate $r$ under a Log-normal($-0.5\sigma^2, \sigma$) distribution (with its mean fixed at 1). 
The function for transforming $\mathcal{R}$ into branch rates $r(\mathcal{R})$ is depicted for \textit{real} (top right), \textit{cat} (bottom left), and \textit{quant} (bottom right).
For visualisation purposes, there are only 10 bins/pieces displayed, however in practice we use $2N-2$ bins for \textit{cat} and 100 pieces for \textit{quant}.
The first and final \textit{quant} pieces are equal to the underlying function (solid lines) however the pieces in between use linear approximations of this function (dashed lines).}
\label{fig:rateparams}
\end{figure}

\clearpage
\subsubsection*{1. Real rates}
The natural (and unabstracted) parameterisation of a substitution rate is a real number $\mathcal{R}_i \in \mathbb{R}, \mathcal{R}_i > 0$ which is equal to the rate itself. 
Under the \textit{real} parameterisation:

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/correlations.pdf}
\caption{\textbf{Traversing likelihood space.}
The z-axes above are the log-likelihoods of the genetic distance $r \times \tau$ between two simulated nucleic acid sequences of length $L$, under the Jukes-Cantor substitution model \cite{jukes1969evolution}. 
Two possible proposals from the current state (white circle) are depicted.
These proposals are generated by the \texttt{RandomWalk} (\texttt{RW}) and \texttt{ConstantDistance} (\texttt{CD}) operators.
In the low signal dataset ($L=0.1$kb), both operators can traverse the likelihood space effectively.
 However, the exact same proposal by \texttt{RandomWalk} incurs a much larger likelihood penalty in the $L=0.5$kb dataset by ``falling off the ridge'', in contrast to \texttt{ConstantDistance} which ``walks along the ridge''.
 This discrepancy is even stronger for larger datasets and thus necessitates the use of operators such as \texttt{ConstantDistance} which account for correlations between branch lengths and rates. }
\label{fig:landscape}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{Figures/clockSD.pdf}
\caption{\textbf{Clock standard deviation scale operators.}
The two operators above propose a clock standard deviation $\sigma \rightarrow \sigma^\prime$. 
Then, either the new quantiles are such that the rates remain constant (``New quantiles'', above) or the new rates are  such that the quantiles remain constant (``New rates''). 
In the \textit{real} parameterisation, these two operators are known as \texttt{Scale} and \texttt{CisScale}, respectively.
Whereas, in \textit{quant}, they are known as \texttt{CisScale} and \texttt{Scale}.    }
\label{fig:clockSDoperators}
\end{figure}

\clearpage
\subsection*{Adaptive operator weighting}
\label{sect:adaptiveSampling}

@@

@@

\begin{table}[h!]
\centering
\caption{\bf{Summary of \texttt{AdaptiveOperatorSampler} operators and their parameters of interest (POI).}}
\begin{tabular}{l l l} 
 Meta-operator & POI & Operators \\
\hline
 \multirow{4}{*}{\texttt{AdaptiveOperatorSampler($\sigma$)}} & \multirow{4}{*}{$\sigma$} & \texttt{CisScale($\sigma, \vec{\mathcal{R}}^{\,}$)} \\ 
 && \texttt{RandomWalk($\sigma$)}  \\
 && \texttt{Scale($\sigma$)}  \\
 && \texttt{SampleFromPrior($\sigma$)}  \\
 \hline
  \multirow{6}{*}{\texttt{AdaptiveOperatorSampler($\vec{\mathcal{R}}^{\,}$)}} & \multirow{6}{*}{$\vec{\mathcal{R}}^{\,}, t$} & \texttt{ConstantDistance($\vec{\mathcal{R}}^{\,}, \mathcal{T}$)}   \\ 
&& \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}$)}  \\
&& \texttt{Scale($\vec{\mathcal{R}}^{\,}$)}   \\
&& \texttt{Interval($\vec{\mathcal{R}}^{\,}$)}   \\
&& \texttt{Swap($\vec{\mathcal{R}}^{\,}$)}  \\
&& \texttt{SampleFromPrior($\vec{\mathcal{R}}^{\,}$)} \\
 \hline
   \multirow{2}{*}{\texttt{AdaptiveOperatorSampler(root)}} & \multirow{2}{*}{$\vec{\mathcal{R}}^{\,}, t$} & \texttt{SimpleDistance($\vec{\mathcal{R}}^{\,}, \mathcal{T}$)}  \\ 
&&  \texttt{SmallPulley($\vec{\mathcal{R}}^{\,}, t$)}  \\
\hline
\hline
 \multirow{7}{*}{\texttt{AdaptiveOperatorSampler(leaf)}} & \multirow{7}{*}{$\vec{\mathcal{R}}^{\,}_\text{leaf}, t$} & \texttt{ConstantDistance($\vec{\mathcal{R}}^{\,}_\text{leaf}, \mathcal{T}$)}   \\ 
&& \texttt{LeafAVMVN($\vec{\mathcal{R}}^{\,}_\text{leaf}$)}  \\
&& \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}_\text{leaf}$)}  \\
&& \texttt{Scale($\vec{\mathcal{R}}^{\,}_\text{leaf}$)}   \\
&& \texttt{Interval($\vec{\mathcal{R}}^{\,}_\text{leaf}$)}   \\
&& \texttt{Swap($\vec{\mathcal{R}}^{\,}_\text{leaf}$)}  \\
&& \texttt{SampleFromPrior($\vec{\mathcal{R}}^{\,}_\text{leaf}$)} \\
\hline
 \multirow{6}{*}{\texttt{AdaptiveOperatorSampler(internal)}} & \multirow{6}{*}{$\vec{\mathcal{R}}^{\,}_\text{int}, t$} & \texttt{ConstantDistance($\vec{\mathcal{R}}^{\,}_\text{int}, \mathcal{T}$)}   \\ 
&& \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}_\text{int}$)}  \\
&& \texttt{Scale($\vec{\mathcal{R}}^{\,}_\text{int}$)}   \\
&& \texttt{Interval($\vec{\mathcal{R}}^{\,}_\text{int}$)}   \\
&& \texttt{Swap($\vec{\mathcal{R}}^{\,}_\text{int}$)}  \\
&& \texttt{SampleFromPrior($\vec{\mathcal{R}}^{\,}_\text{int}$)} \\
\hline
\hline
 \multirow{2}{*}{\texttt{AdaptiveOperatorSampler(NER)}} & \multirow{2}{*}{$\mathcal{T}$} & NER$\{  \}$ \\ 
 && NER$\{ \mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE} \}$ \\
\end{tabular}
\begin{flushleft}
Different operators are applicable to different substitution rate parameterisations (\textbf{Table \ref{table:kernels}}).
Nodes are broken down into regions to enable operators to be weighted according to their dimension.  
\texttt{AdaptiveOperatorSampler(root)} applies the root-targeting constant distance operators only \cite{zhang2020improving} while \texttt{AdaptiveOperatorSampler($\vec{\mathcal{R}}^{\,}$)} targets all rates and all nodes heights $t$. 
Nodes are further broken down into leaf rate $\vec{\mathcal{R}}^{\,}_\text{leaf}$ and internal node rate $\vec{\mathcal{R}}^{\,}_\text{int}$ operators.
This facilitates suitable weighting of the \texttt{LeafAVMVN} operator, which is only applicable to leaf nodes (\textbf{\nameref{AVMVN_sect}}).
In this setup, the \texttt{RandomWalk($x$)}, \texttt{Scale($x$)}, and \texttt{SampleFromPrior($x$)} operators apply to the corresponding set of branch rates $x$, whereas \texttt{ConstantDistance($x, \mathcal{T}$)} is only applicable to internal nodes which have at least one child of type $x \in \{\vec{\mathcal{R}}^{\,}_\text{leaf}, \vec{\mathcal{R}}^{\,}_\text{int}\}$.
Finally, the Robinson-Foulds distance between trees before and after every proposal accept is used to train the weights behind \texttt{AdaptiveOperatorSampler(NER)} (see \textbf{\nameref{sect:NER}}). In the special case of NER proposals, the \texttt{RF} distance is always equal to 1. 
\end{flushleft}
\label{table:adaptiveSampling}
\end{table}

\clearpage
\subsection*{Bactrian proposal kernel} \label{sect:randomwalks}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/bactrian.pdf}
\caption{\textbf{The Bactrian proposal kernel.} The step size made under a Bactrian proposal kernel is equal to $s \Sigma$ where $\Sigma$ is drawn from the above distribution and $s$ is tunable.   }
\label{fig:bactrian}
\end{figure}

@@

\clearpage
\subsection*{Narrow Exchange Rate} \label{sect:NER}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/NarrowExchange.pdf}
\caption{\textbf{Depiction of \texttt{NarrowExchange} and \texttt{NarrowExchangeRate} operators.} Proposals are denoted by $\mathcal{T} \rightarrow \mathcal{T}^\prime$. The vertical axes correspond to node heights $t$. In the bottom figure, branch rates $r$ are indicated by line width and therefore genetic distances are equal to the width of each branch multiplied by its length. 
In this example, the $\mathcal{D}_{AE}$ and $\mathcal{D}_{CE}$ constraints are satisfied.
%RB: at this stage, D_AE and D_CE constraints have not been defined yet. Remove or reformulate or move to after definition?
}
\label{fig:narrowexchange}
\end{figure}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/acceptanceRates.pdf}
\caption{\textbf{Screening of NER and NERw variants by acceptance rate.} Top left: comparison of NER variants with the null operator $\text{NER}\{\}$ = \texttt{NarrowExchange}. 
Each operator is represented by a single point, uniquely encoded by the point stylings. 
The number of times each operator is proposed and accepted is compared with that of $\text{NER}\{\}$, and one-sided z-tests are performed to assess the statistical significance between the two acceptance rates ($p = 0.001$).  This process is repeated across $300$ simulated datasets. The axes of each plot are the proportion of these $300$ simulations for which there is evidence that the operator is significantly better than $\text{NER}\{\}$ (x-axis) or worse than $\text{NER}\{\}$ (y-axis). Top right: comparison of NER and NERw acceptance rates. Each point is one NER/NERw variant from a single simulation. Bottom: relationship between the acceptance rates $\alpha$ of $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ and $\text{NER}\{ \}$ with the clock model standard deviation $\sigma$ and the number of sites $L$.  Each point is a single simulation. }
\label{fig:acceptanceRateScreening}
\end{figure}

@@

To prevent the learned weight behind \texttt{LeafAVMVN} from dominating the \texttt{AdaptiveOperatorSampler} weighting scheme and therefore inhibiting the mixing of internal node rates, we 
introduce the \texttt{AdaptiveOperatorSampler(leaf)} and \texttt{AdaptiveOperatorSampler(internal)} meta-operators which operate exclusively on leaf node rates $\vec{\mathcal{R}}^{\,}_\text{leaf}$ and internal node rates $\vec{\mathcal{R}}^{\,}_\text{int}$ respectively  (\textbf{Table \ref{table:AVMVNoperators}}). 
The former employs the \texttt{LeafAVMVN} operator and learns its weight during MCMC (after providing it sufficient time to learn $\Sigma_N$).

\clearpage
\subsection*{Model specification and MCMC settings} \label{sect:methods}

@@

\begin{table}[h!]
\centering
\caption{\bf{Summary of clock model operators introduced throughout this article. }}
\begin{tabular}{l p{4cm} l} 
 Operator & Description & Parameters  \\
  \hline
 \texttt{AdaptiveOperatorSampler} & Samples sub-operators proportionally to their weights, which are learned (see \nameref{sect:adaptiveSampling}). & $\vec{\mathcal{R}}^{\,}, \sigma, \mathcal{T}$ \\
  \hline
 \texttt{SampleFromPrior} & Resamples a random number of elements from their prior (see \nameref{sect:adaptiveSampling}). & $\vec{\mathcal{R}}^{\,}, \sigma$ \\
  \hline
 \texttt{NarrowExchangeRate} & Moves a branch and recomputes branch rates so that their genetic distances are constant (see \nameref{sect:NER}). & $\vec{\mathcal{R}}^{\,}, \mathcal{T}$\\
  \hline
 \texttt{LeafAVMVN}  & Proposes new rates for all leaves in one move (see \nameref{AVMVN_sect}) \cite{baele2017adaptive}. & $\vec{\mathcal{R}}^{\,}$ \\
\end{tabular}
\begin{flushleft}
Pre-existing clock model operators are summarised in \textbf{Table \ref{table:kernels}}
\end{flushleft}
\label{table:newOperators}
\end{table}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/tournament.pdf}
\caption{\textbf{Protocol for optimising clock model methodologies.} Each area (detailed in \textbf{\nameref{sect:models}}) is optimised sequentially, and the best setting from each step is used when optimising the following step.}
\label{fig:tournament}
\end{figure}

@@

Methodologies were benchmarked using one simulated and eight empirical datasets.
The latter were compiled by Lanfear as ``benchmark alignments'' (\textbf{Table \ref{table:datasets}}) \cite{lanfear2019Github, lanfear2016partitionfinder}.
Each methodology was benchmarked for million-states-per-hour using the Intel Xeon Gold 6138 CPU (2.00 GHz).
These terms were multiplied by the ESS-per-state across 20 replicates on the New Zealand eScience Infrastructure (NeSI) cluster to compute the total ESS/hr of each dataset under each setting.
All methodologies used identical models and operator configurations, except where a difference is specified.

%All 25 datasets presented in \textbf{Table \ref{table:datasets}} are benchmarked together $1 \times$ each at the end of the protocol. Datasets and BEAST 2 .xml templates are provided in the GitHub repository accompanying this article.

@@

\clearpage
\subsection*{Round 1: A simple operator-weight learning algorithm greatly improved performance}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{benchmarking/benchmarkingVM/ESS_round1_real.pdf}
\caption{\textbf{Round 1: benchmarking the \texttt{AdaptiveOperatorSampler} operator.} Top left, top right, bottom left: each plot compares the ESS/hr ($\pm 1$ standard error) across two operator configurations.
Bottom right: the effect of sequence length $L$ on operator weights learned by \texttt{AdaptiveOperatorSampler}. Both sets of observations are fit by logistic regression models.
 The benchmark datasets are displayed in \textbf{Table \ref{table:datasets}}. The \emph{cat} and \emph{quant} settings are evaluated in \textbf{\nameref{sect:SR_appendix}}.  }
\label{fig:round1Results}
\end{figure}

\clearpage
\subsection*{Round 2: The \textit{real} parameterisation yielded the fastest mixing }

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{benchmarking/benchmarkingVM/ESS_round2.pdf}
\caption{\textbf{Round 2: benchmarking substitution rate parameterisations.} Top left, top right, bottom left: the adapt (\textit{real}), adapt (\textit{cat}), and adapt (\textit{quant}) configurations were compared. Bottom right: comparison of the mean tip substitution rate ESS/hr as a function of alignment length $L$. }
\label{fig:round2Results}
\end{figure}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{benchmarking/benchmarkingVM/ESS_timesettings.pdf}
\caption{\textbf{Comparison of runtimes across methodologies.} 
The computational time required for a setting to sample a single state is divided by that of the nocons (\textit{cat}) configuration.
The geometric mean under each configuration, averaged across all 9 datasets, is displayed as a horizontal bar.}
\label{fig:round2Resultsb}
\end{figure}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{benchmarking/benchmarkingVM/ESS_round3.pdf}
\caption{\textbf{Round 3: benchmarking the Bactrian kernel.}
The ESS/hr ($\pm 1$ s.e.) under the Bactrian configuration, divided by that under the uniform kernel, is shown in the y-axis for each dataset and relevant parameter.
Horizontal bars show the geometric mean under each parameter.
 }
\label{fig:round3Results}
\end{figure}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{benchmarking/benchmarkingVM/ESS_round4.pdf}
\includegraphics[width=\textwidth]{Figures/broughton.pdf}
\caption{\textbf{Round 4: benchmarking the NER operators.} 
Top: the learned weights (left) behind the two NER operators ($\text{NER} \{\}$ and $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$), and the relative difference between their acceptance rates $\alpha$ (right), are presented as functions of sequence length.
Logistic and logarithmic regression models are shown, respectively.
Bottom: maximum clade credibility tree of the bony fish dataset by Broughton et al. 2013 \cite{Broughton_2013}.
This alignment received the strongest boost from NER, likely due to its high topological uncertainty and branch rate variance.  
Branches are coloured by substitution rate, the y-axis shows time units, and internal nodes are labelled with posterior clade support.  
Tree visualised using UglyTrees \cite{uglytrees}.}
\label{fig:round4Results}
\end{figure}

\clearpage
\subsection*{Round 5: The AVMVN leaf rate operator was computationally demanding and improved mixing very slightly}

@@

\newpage
\section*{Conclusion}

@@

\section*{Software availability}

>>>>

\subsection*{Branch rate parameterisations}
\label{sect:rateparams}

@@

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{Figures/rateparameterisation.pdf}
\caption{\textbf{Branch rate parameterisations.} Top left: the prior density of a branch rate $r$ under a Log-normal($-0.5\sigma^2, \sigma$) distribution (with its mean fixed at 1). 
The function for transforming $\mathcal{R}$ into branch rates $r(\mathcal{R})$ is depicted for \textit{real} (top right), \textit{cat} (bottom left), and \textit{quant} (bottom right).
For visualisation purposes, there are only 10 bins/pieces displayed, however in practice we use $2N-2$ bins for \textit{cat} and 100 pieces for \textit{quant}.
The first and final \textit{quant} pieces are equal to the underlying function (solid lines) however the pieces in between use linear approximations of this function (dashed lines).}
\label{fig:rateparams}
\end{figure}

\subsubsection*{1. Real rates}
The natural (and unabstracted) parameterisation of a substitution rate is a real number $\mathcal{R}_i \in \mathbb{R}, \mathcal{R}_i > 0$ which is equal to the rate itself. 
Under the \textit{real} parameterisation:

@@

@@

\begin{figure}[!h]
\centering
%\includegraphics[width=\textwidth]{Figures/clockSD.pdf}
\caption{\textbf{Clock standard deviation scale operators.}
The two operators above propose a clock standard deviation $\sigma \rightarrow \sigma^\prime$. 
Then, either the new quantiles are such that the rates remain constant (``New quantiles'', above) or the new rates are  such that the quantiles remain constant (``New rates''). 
In the \textit{real} parameterisation, these two operators are known as \texttt{Scale} and \texttt{CisScale}, respectively.
Whereas, in \textit{quant}, they are known as \texttt{CisScale} and \texttt{Scale}.    }
\label{fig:clockSDoperators}
\end{figure}

\subsection*{Adaptive operator weighting}
\label{sect:adaptiveSampling}

@@

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{Figures/correlations.pdf}
\caption{\textbf{Traversing likelihood space.}
The z-axes above are the log-likelihoods of the genetic distance $r \times \tau$ between two simulated nucleic acid sequences of length $L$, under the Jukes-Cantor substitution model \cite{jukes1969evolution}. 
Two possible proposals from the current state (white circle) are depicted.
These proposals are generated by the \texttt{RandomWalk} (\texttt{RW}) and \texttt{ConstantDistance} (\texttt{CD}) operators.
In the low signal dataset ($L=0.1$kb), both operators can traverse the likelihood space effectively.
 However, the exact same proposal by \texttt{RandomWalk} incurs a much larger likelihood penalty in the $L=0.5$kb dataset by ``falling off the ridge'', in contrast to \texttt{ConstantDistance} which ``walks along the ridge''.
 This discrepancy is even stronger for larger datasets and thus necessitates the use of operators such as \texttt{ConstantDistance} which account for correlations between branch lengths and rates. }
\label{fig:landscape}
\end{figure}

@@

\begin{table}[h!]
\centering
\caption{\bf{Summary of \texttt{AdaptiveOperatorSampler} operators and their parameters of interest (POI).}}
\begin{tabular}{l l l} 
 Meta-operator & POI & Operators \\
\hline
 \multirow{4}{*}{\texttt{AdaptiveOperatorSampler($\sigma$)}} & \multirow{4}{*}{$\sigma$} & \texttt{CisScale($\sigma, \vec{\mathcal{R}}^{\,}$)} \\ 
 && \texttt{RandomWalk($\sigma$)}  \\
 && \texttt{Scale($\sigma$)}  \\
 && \texttt{SampleFromPrior($\sigma$)}  \\
 \hline
  \multirow{6}{*}{\texttt{AdaptiveOperatorSampler($\vec{\mathcal{R}}^{\,}$)}} & \multirow{6}{*}{$\vec{\mathcal{R}}^{\,}, t$} & \texttt{ConstantDistance($\vec{\mathcal{R}}^{\,}, t$)}   \\ 
&& \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}$)}  \\
&& \texttt{Scale($\vec{\mathcal{R}}^{\,}$)}   \\
&& \texttt{Interval($\vec{\mathcal{R}}^{\,}$)}   \\
&& \texttt{Swap($\vec{\mathcal{R}}^{\,}$)}  \\
&& \texttt{SampleFromPrior($\vec{\mathcal{R}}^{\,}$)} \\
 \hline
   \multirow{2}{*}{\texttt{AdaptiveOperatorSampler(root)}} & \multirow{2}{*}{$\vec{\mathcal{R}}^{\,}, t$} & \texttt{SimpleDistance($\vec{\mathcal{R}}^{\,}, t$)}  \\ 
&&  \texttt{SmallPulley($\vec{\mathcal{R}}^{\,}, t$)}  \\
\hline
\hline
 \multirow{7}{*}{\texttt{AdaptiveOperatorSampler(leaf)}} & \multirow{7}{*}{$\vec{\mathcal{R}}^{\,}_\text{leaf}, t$} & \texttt{ConstantDistance($\vec{\mathcal{R}}^{\,}_\text{leaf}, t$)}   \\ 
&& \texttt{LeafAVMVN($\vec{\mathcal{R}}^{\,}_\text{leaf}$)}  \\
&& \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}_\text{leaf}$)}  \\
&& \texttt{Scale($\vec{\mathcal{R}}^{\,}_\text{leaf}$)}   \\
&& \texttt{Interval($\vec{\mathcal{R}}^{\,}_\text{leaf}$)}   \\
&& \texttt{Swap($\vec{\mathcal{R}}^{\,}_\text{leaf}$)}  \\
&& \texttt{SampleFromPrior($\vec{\mathcal{R}}^{\,}_\text{leaf}$)} \\
\hline
 \multirow{6}{*}{\texttt{AdaptiveOperatorSampler(internal)}} & \multirow{6}{*}{$\vec{\mathcal{R}}^{\,}_\text{int}, t$} & \texttt{ConstantDistance($\vec{\mathcal{R}}^{\,}_\text{int}, t$)}   \\ 
&& \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}_\text{int}$)}  \\
&& \texttt{Scale($\vec{\mathcal{R}}^{\,}_\text{int}$)}   \\
&& \texttt{Interval($\vec{\mathcal{R}}^{\,}_\text{int}$)}   \\
&& \texttt{Swap($\vec{\mathcal{R}}^{\,}_\text{int}$)}  \\
&& \texttt{SampleFromPrior($\vec{\mathcal{R}}^{\,}_\text{int}$)} \\
\hline
\hline
 \multirow{2}{*}{\texttt{AdaptiveOperatorSampler(NER)}} & \multirow{2}{*}{$\mathcal{T}$} & NER$\{  \}$ \\ 
 && NER$\{ \mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE} \}$ \\
\end{tabular}
\begin{flushleft}
Different operators are applicable to different substitution rate parameterisations (\textbf{Table \ref{table:kernels}}).
Nodes are broken down into regions to enable operators to be weighted according to their dimension.  
\texttt{AdaptiveOperatorSampler(root)} applies the root-targeting constant distance operators only \cite{zhang2020improving} while \texttt{AdaptiveOperatorSampler($\vec{\mathcal{R}}^{\,}$)} targets all rates and all nodes heights $t$. 
Nodes are further broken down into leaf rate $\vec{\mathcal{R}}^{\,}_\text{leaf}$ and internal node rate $\vec{\mathcal{R}}^{\,}_\text{int}$ operators.
This facilitates suitable weighting of the \texttt{LeafAVMVN} operator, which is only applicable to leaf nodes (\textbf{\nameref{AVMVN_sect}}).
In this setup, the \texttt{RandomWalk($x$)}, \texttt{Scale($x$)}, and \texttt{SampleFromPrior($x$)} operators apply to the corresponding set of branch rates $x$, whereas \texttt{ConstantDistance($x, t$)} is only applicable to internal nodes which have at least one child of type $x \in \{\vec{\mathcal{R}}^{\,}_\text{leaf}, \vec{\mathcal{R}}^{\,}_\text{int}\}$.
Finally, the Robinson-Foulds distance between trees before and after every proposal accept is used to train the weights behind \texttt{AdaptiveOperatorSampler(NER)} (see \textbf{\nameref{sect:NER}}). In the special case of NER proposals, the \texttt{RF} distance is always equal to 1. 
\end{flushleft}
\label{table:adaptiveSampling}
\end{table}

\subsection*{Bactrian proposal kernel} \label{sect:randomwalks}

@@

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{Figures/bactrian.pdf}
\caption{\textbf{The Bactrian proposal kernel.} The step size made under a Bactrian proposal kernel is equal to $s \Sigma$ where $\Sigma$ is drawn from the above distribution and $s$ is tunable.   }
\label{fig:bactrian}
\end{figure}

@@

\subsection*{Narrow Exchange Rate} \label{sect:NER}

@@

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{Figures/NarrowExchange.pdf}
\caption{\textbf{Depiction of \texttt{NarrowExchange} and \texttt{NarrowExchangeRate} operators.} Proposals are denoted by $\mathcal{T} \rightarrow \mathcal{T}^\prime$. The vertical axes correspond to node heights $t$. In the bottom figure, branch rates $r$ are indicated by line width and therefore genetic distances are equal to the width of each branch multiplied by its length. 
In this example, the $\mathcal{D}_{AE}$ and $\mathcal{D}_{CE}$ constraints are satisfied.
%RB: at this stage, D_AE and D_CE constraints have not been defined yet. Remove or reformulate or move to after definition?
}
\label{fig:narrowexchange}
\end{figure}

@@

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{Figures/acceptanceRates.pdf}
\caption{\textbf{Screening of NER and NERw variants by acceptance rate.} Top left: comparison of NER variants with the null operator $\text{NER}\{\}$ = \texttt{NarrowExchange}. 
Each operator is represented by a single point, uniquely encoded by the point stylings. 
The number of times each operator is proposed and accepted is compared with that of $\text{NER}\{\}$, and one-sided z-tests are performed to assess the statistical significance between the two acceptance rates ($p = 0.001$).  This process is repeated across $300$ simulated datasets. The axes of each plot are the proportion of these $300$ simulations for which there is evidence that the operator is significantly better than $\text{NER}\{\}$ (x-axis) or worse than $\text{NER}\{\}$ (y-axis). Top right: comparison of NER and NERw acceptance rates. Each point is one NER/NERw variant from a single simulation. Bottom: relationship between the acceptance rates $\alpha$ of $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ and $\text{NER}\{ \}$ with the clock model standard deviation $\sigma$ and the number of sites $L$.  Each point is a single simulation. }
\label{fig:acceptanceRateScreening}
\end{figure}

@@

To prevent the learned weight behind \texttt{LeafAVMVN} from dominating the \texttt{AdaptiveOperatorSampler} weighting scheme and therefore inhibiting the mixing of internal node rates, we 
introduce the \texttt{AdaptiveOperatorSampler(leaf)} and \texttt{AdaptiveOperatorSampler(internal)} meta-operators which operate exclusively on leaf node rates $\vec{\mathcal{R}}^{\,}_\text{leaf}$ and internal node rates $\vec{\mathcal{R}}^{\,}_\text{int}$ respectively  (\textbf{Table \ref{table:adaptiveSampling}}). 
The former employs the \texttt{LeafAVMVN} operator and learns its weight during MCMC (after providing it sufficient time to learn $\Sigma_N$).

\subsection*{Model specification and MCMC settings} \label{sect:methods}

@@

\begin{table}[h!]
\centering
\caption{\bf{Summary of clock model operators introduced throughout this article. }}
\begin{tabular}{l p{4cm} l} 
 Operator & Description & Parameters  \\
  \hline
 \texttt{AdaptiveOperatorSampler} & Samples sub-operators proportionally to their weights, which are learned (see \nameref{sect:adaptiveSampling}). & $\vec{\mathcal{R}}^{\,}, \sigma, t \mathcal{T}$ \\
  \hline
 \texttt{SampleFromPrior} & Resamples a random number of elements from their prior (see \nameref{sect:adaptiveSampling}). & $\vec{\mathcal{R}}^{\,}, \sigma$ \\
  \hline
 \texttt{NarrowExchangeRate} & Moves a branch and recomputes branch rates so that their genetic distances are constant (see \nameref{sect:NER}). & $\vec{\mathcal{R}}^{\,}, \mathcal{T}$\\
  \hline
 \texttt{LeafAVMVN}  & Proposes new rates for all leaves in one move (see \nameref{AVMVN_sect}) \cite{baele2017adaptive}. & $\vec{\mathcal{R}}^{\,}$ \\
\end{tabular}
\begin{flushleft}
Pre-existing clock model operators are summarised in \textbf{Table \ref{table:kernels}}. $t$ denotes node heights while $\mathcal{T}$ denotes the whole tree, including topology.
\end{flushleft}
\label{table:newOperators}
\end{table}

@@

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{Figures/tournament.pdf}
\caption{\textbf{Protocol for optimising clock model methodologies.} Each area (detailed in \textbf{\nameref{sect:models}}) is optimised sequentially, and the best setting from each step is used when optimising the following step.}
\label{fig:tournament}
\end{figure}

@@

Methodologies were benchmarked using one simulated and eight empirical datasets \cite{Ran_2018,Dornburg_2012,Cognato_2001,Sauquet_2011,Broughton_2013,Kawahara_2013,Rightmyer_2013,Moyle_2016}.
The latter were compiled by Lanfear as ``benchmark alignments'' (\textbf{Table \ref{table:datasets}}) \cite{lanfear2019Github, lanfear2016partitionfinder}.
Each methodology was benchmarked for million-states-per-hour using the Intel Xeon Gold 6138 CPU (2.00 GHz).
These terms were multiplied by the ESS-per-state across 20 replicates on the New Zealand eScience Infrastructure (NeSI) cluster to compute the total ESS/hr of each dataset under each setting.
All methodologies used identical models and operator configurations, except where a difference is specified.

@@

\subsection*{Round 1: A simple operator-weight learning algorithm greatly improved performance}

@@

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{benchmarking/benchmarkingVM/ESS_round1_real.pdf}
\caption{\textbf{Round 1: benchmarking the \texttt{AdaptiveOperatorSampler} operator.} Top left, top right, bottom left: each plot compares the ESS/hr ($\pm 1$ standard error) across two operator configurations.
Bottom right: the effect of sequence length $L$ on operator weights learned by \texttt{AdaptiveOperatorSampler}. Both sets of observations are fit by logistic regression models.
 The benchmark datasets are displayed in \textbf{Table \ref{table:datasets}}. The \emph{cat} and \emph{quant} settings are evaluated in \textbf{\nameref{sect:SR_appendix}}.  }
\label{fig:round1Results}
\end{figure}

\subsection*{Round 2: The \textit{real} parameterisation yielded the fastest mixing }

@@

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{benchmarking/benchmarkingVM/ESS_round2.pdf}
\caption{\textbf{Round 2: benchmarking substitution rate parameterisations.} Top left, top right, bottom left: the adapt (\textit{real}), adapt (\textit{cat}), and adapt (\textit{quant}) configurations were compared. Bottom right: comparison of the mean tip substitution rate ESS/hr as a function of alignment length $L$. }
\label{fig:round2Results}
\end{figure}

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{benchmarking/benchmarkingVM/ESS_timesettings.pdf}
\caption{\textbf{Comparison of runtimes across methodologies.} 
The computational time required for a setting to sample a single state is divided by that of the nocons (\textit{cat}) configuration.
The geometric mean under each configuration, averaged across all 9 datasets, is displayed as a horizontal bar.}
\label{fig:round2Resultsb}
\end{figure}

@@

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{benchmarking/benchmarkingVM/ESS_round3.pdf}
\caption{\textbf{Round 3: benchmarking the Bactrian kernel.}
The ESS/hr ($\pm 1$ s.e.) under the Bactrian configuration, divided by that under the uniform kernel, is shown in the y-axis for each dataset and relevant parameter.
Horizontal bars show the geometric mean under each parameter.
 }
\label{fig:round3Results}
\end{figure}

@@

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{Figures/Fig12.pdf}
\caption{\textbf{Round 4: benchmarking the NER operators.} 
Top: the learned weights (left) behind the two NER operators ($\text{NER} \{\}$ and $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$), and the relative difference between their acceptance rates $\alpha$ (right), are presented as functions of sequence length.
Logistic and logarithmic regression models are shown, respectively.
Bottom: maximum clade credibility tree of the bony fish dataset by Broughton et al. 2013 \cite{Broughton_2013}.
This alignment received the strongest boost from NER, likely due to its high topological uncertainty and branch rate variance.  
Branches are coloured by substitution rate, the y-axis shows time units, and internal nodes are labelled with posterior clade support.  
Tree visualised using UglyTrees \cite{uglytrees}.}
\label{fig:round4Results}
\end{figure}

\subsection*{Round 5: The AVMVN leaf rate operator was computationally demanding and improved mixing very slightly}

@@

\section*{Conclusion}

@@
